---
title: "Hw5"
format: html
editor: visual
---

## Read in the Data and Modify

Using code from the provided file to import and merge both datasets.
```{r}
d1 <- read.table("student-mat.csv",sep=";",header=TRUE)
d2 <- read.table("student-por.csv",sep=";",header=TRUE)

d3 <- merge(d1,d2,by=c("school","sex","age","address","famsize","Pstatus","Medu","Fedu","Mjob","Fjob","reason","nursery","internet"))
d3
```

Doing the same process but using tidyverse functions read_csv2 (semicolon seperated file) and inner_join. This leads to differences in how the rows are sorted, with the output data above having nicely ordered row arrangements for variables such as school, sex, and age. By comparison the output using tidyverse below is data that is not ordered with any variable other than school. Also the integer output variables above are instead of the data type double in the data output below, and for variables not included in the join parameters, two different columns are created in the dataset, differentiated by .x and .y appended to the end.
```{r}
library(tidyverse)
school1 <- read_csv2("student-mat.csv")
school2 <- read_csv2("student-por.csv")

school <- inner_join(school1,school2,by=c("school","sex","age","address","famsize","Pstatus","Medu","Fedu","Mjob","Fjob","reason","nursery","internet"))
school
```

Using inner_join to merge all variables other than G1, G2, G3, paid, and absences.
```{r}
school <- inner_join(school1,school2,by = names(school1)[!names(school1) %in% c("G1","G2", "G3", "paid", "absences")])
school
```

Picking 4 categorical variables of interest (school, sex, address, and Pstatus) and converting these to factor variables using mutate.
```{r}
school <- school |> mutate(across(c("school", "sex", "address", "Pstatus"), as.factor))
school
```

## Summarize the Data

Checking how the data is stored and documenting all missing values in the data.
```{r}
str(school)
colSums(is.na(school))
```

Creating a one way, two way, and three way table for our previously chosen factor variables. 
For table 1: 146 refers to the total number of males in this data set.
For table 2: 134 refers to the subset of males that attend the school Gabriel Pereira.
For table 3: 25 refers to number of male students that attend Gabriel Pereira who live in a rural area.
```{r}
table(school$sex)
table(school$sex, school$school)
table(school$sex, school$school, school$address)
```

Two different ways to create a conditional two-way table, one using filter and the other subsetting a three-way table.
```{r}
schoolGP <- school |> filter(school == "GP")
table(schoolGP$sex, schoolGP$Pstatus)

threewayschool <- table(school$sex, school$school, school$Pstatus)
threewayschool[,1,]
```

Creating a two way contingency table using group_by and summarize. Afterwards the result is stretched using pivot_wider.
```{r}
twowayschool <- school |> group_by(school, sex) |> summarize(count = n()) 
twowayschool |> pivot_wider(names_from = school, values_from = count)
```

Creating a stacked bar graph and side-by-side bar graph.
```{r}
school |> ggplot(aes(x = sex)) + geom_bar(aes(fill = school)) + labs(title = "Student Population of Two Different Schools") + xlab("Sex") + ylab("Count")

school |> ggplot(aes(x = sex, fill = school)) + geom_bar(position = "dodge") + labs(title = "Student Population of Two Different Schools") + xlab("Sex") + ylab("Count")
```

## Numeric Variables

Finding measures of center and spread for age, absences, and one G3 variable. Repeating to subset the data. This table reveals that the measures of center and spread for all of these variables match very closely between the overall data set and when subsetted by looking at only the Gabriel Pereira school. This makes sense as we observed previously that GP has the greatest share of the student population by far.
```{r}
school |> summarize(across(c("age", "absences.x", "absences.y", "G3.x"), list("mean" = mean, "median" = median, "SD" = sd), .names = "{.fn}_{.col}"))

school |> filter(school == "GP") |> summarize(across(c("age", "absences.x", "absences.y", "G3.x"), list("mean" = mean, "median" = median, "SD" = sd), .names = "{.fn}_{.col}"))
```

Finding measures of center and spread across a single grouping variable. One interesting thing revealed through the table is that it appears that female students have slightly more absences on average than male students, despite having the same median, and the number of absences across each population is more spread out for women than for men.
```{r}
school |> group_by(sex) |> summarize(across(c("age", "absences.x", "absences.y", "G3.x"), list("mean" = mean, "median" = median, "SD" = sd), .names = "{.fn}_{.col}"))
```

Finding measures of center and spread across two grouping variables. One interesting thing revealed by this table is that for both sexes, students at the Mousinho da Silveira school are slightly older than students at Gabriel Pereira, yet their G3 grades are lower.
```{r}
school |> group_by(sex, school) |> summarize(across(c("age", "absences.x", "absences.y", "G3.x"), list("mean" = mean, "median" = median, "SD" = sd), .names = "{.fn}_{.col}"))
```

Creating a correlation matrix between all of the numeric variables included in the school data object. Several observations can be gleaned from this. Some are expected, such as absences trending negatively with all grade variables, whereas some I did not think to expect, such as age trending negatively with most of the grade variables.
```{r}
cor(school[,c("age", "absences.x", "absences.y", "G1.x", "G1.y", "G2.x", "G2.y", "G3.x", "G3.y")], use = "complete.obs")
```

Creating a histogram, kernel density plot, and boxplot for two numeric variables, G3.x and G3.y, across the categorical variable school. Based on these plots, it seems that grades in the G3.y group were skewed slightly higher than those in the G3.x group, which have lower values. 
```{r}
schoolplot <- school |> select(school, G3.x, G3.y) |> pivot_longer(cols = c(G3.x, G3.y), names_to = "variable", values_to = "values")
schoolplot |> ggplot(aes(x = values, fill = interaction(school, variable))) + geom_histogram(alpha = 0.8) + labs(title = "G3 Grades by School", x = "Grade", y = "Count")

schoolplot |> ggplot(aes(x = values, fill = interaction(school, variable))) + geom_density(alpha = 0.5) + labs(title = "G3 Grades by School", x = "Grade", y = "Proportion")

schoolplot |> ggplot(aes(x = values, y = variable, color = school)) + geom_boxplot() + labs(title = "G3 Grades by School", x = "Grade", y = "Class")
```

Two scatter plots relating a G3 variable to other numeric variables, one is G3.x vs absences.x, the other G3.y vs absences.y. Each plot is colored by school. From this we can also see the visualization of the trend noted previously, where the .y class (class 2 as named for convenience) appears to have better grades for G3 than the .x class (or class 1). We can also observe in this plot that there are fewer students with more than 20 absences in class 2 compared to class 1, which might help explain the difference
```{r}
school |> ggplot(aes(x = absences.x, y = G3.x, color = school)) + geom_point() + labs(title = "G3 Grades vs Number of Absences at each school (Class 1)", x = "Absences", y = "Grade")

school |> ggplot(aes(x = absences.y, y = G3.y, color = school)) + geom_point() + labs(title = "G3 Grades vs Number of Absences at each school (Class 2)", x = "Absences", y = "Grade")
```
